Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-20 talk/iwtc11/paper.tex
go over the allocation-removal section and replace "virtual" by "allocation-removed". Also some small simplifications.

\subsection{Allocation Removals}
By using escape analysis it is possible to identify objects that are
allocated within the loop but never escape it. That is 
short lived objects with no references outside the loop. This
is performed by processing the operation in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get} and \lstinline{set}) on the removed objects are also
removed and the optimizer needs to keep track of the value of all
used attributes of the object.

Consider again the original unoptimized trace of
Figure~\ref{fig:peeled-trace}. Line 10 contains the first
allocation. It is removed and $p_5$ is marked as virtual. This means
that it refers to an virtual object that has not yet been
(and might never be) allocated. Line 12 sets the \lstinline{intval}
attribute of $p_5$. This operation is also removed and the optimizer
registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.

When the optimizer reaches line 13 it needs to construct the
arguments of the \lstinline{jump} operation, which contains the virtual
reference $p_5$. This can be achieved by exploding $p_5$ into it's
attributes. In this case there is only one attribute and it's value is
$i_4$, which means that $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each virtual in the jump arguments is exploded into a
vector of variables containing the values of all registered attributes. If some
of the attributes are themselves virtuals they are recursively exploded
to make the vector contain only non-virtual variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every non-
virtual variable of the jump arguments is exploded into a vector. This will
be a vector containing the original variable only. To summarize, for
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
        H^{\left(k\right)} & \text{if $J_k$ is virtual}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all non virtual
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the preamble. This
explosion can't be repeated when constructing the arguments of the
jump of the peeled loop as it has to match the first jump. This means that
the objects that was passed as pointers (non virtuals) from the first
iteration to the second (from preamble to peeled loop) also has to be
passed as pointers from the second iteration to the third (from peeled
loop to peeled loop). If one of these objects are virtual 
at the end of the peeled loop they need to be allocated right
before the jump. With the simple objects considered in this paper,
that is not a problem. However in more complicated interpreters such
an allocation might, in combination with other optimizations, lead
to additional variables from the preamble being imported into
the second. This extends both $\hat J$ and $\hat K$, which means that
some care has to be taken, when implementing this, to allow $\hat J$ to
grow while inlining it into $\hat K$. XXX: Maybe we can skip this?

>>>>

\subsection{Allocation Removals}
PyPy's allocation removal optimization \cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. Those objects have to be allocated in the loop, but no outside
object ever gets a reference short lived objects with no references outside the
loop. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

Consider again the original unoptimized trace of
Figure~\ref{fig:peeled-trace}. Line 10 contains the first
allocation. It is removed and $p_5$ is marked as allocation-removed. This means
that it refers to an object that has not yet been
(and might never be) allocated. Line 12 sets the \lstinline{intval}
attribute of $p_5$. This operation is also removed and the optimizer
registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.

When the optimizer reaches line 13 it needs to construct the
arguments of the \lstinline{jump} operation, which contains the
reference to the allocation-removed object in $p_5$. This can be achieved by
exploding $p_5$ into the fields of the allocation-removed object.
In this case there is only one such field and its value is
$i_4$, which means that $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
fields\footnote{This is sometimes called \emph{scalar replacement}. XXX check
whether that's true}. If some of the fields are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the fields in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the preamble. This
explosion can't be repeated when constructing the arguments of the
jump of the peeled loop as it has to match the first jump. This means that
the objects that was passed as pointers (non virtuals) from the first
iteration to the second (from preamble to peeled loop) also has to be
passed as pointers from the second iteration to the third (from peeled
loop to peeled loop). If one of these objects are allocation-removed
at the end of the peeled loop they need to be allocated right
before the jump. With the simple objects considered in this paper,
that is not a problem. However in more complicated interpreters such
an allocation might, in combination with other optimizations, lead
to additional variables from the preamble being imported into
the second. This extends both $\hat J$ and $\hat K$, which means that
some care has to be taken, when implementing this, to allow $\hat J$ to
grow while inlining it into $\hat K$. XXX: Maybe we can skip this?
